Bacterial Microbiome and Obesity Meta-analysis Supplemental

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016.Rmd
Light editing of text

## Introduction
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014]. This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014]. Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013]. Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016]. If the microbiome does affect obesity status, then manipulating the microbiome could have a significant role in the future treatment of obesity and in helping to stem the current epidemic.

@@

One critique of these re-analysis studies is that they did not aggregate the results across studies to increase the effective sample size. It is possible that there are small associations within each study that would not be significant, because the studies lacked sufficient statistical power. The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously. Our goal was to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status.

## Methods
***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1]. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified nine additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity. The nine studies are summarized in Tables 1 and 2.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. We also measured the Spearman correlation coefficient between each metric and the actual BMI value. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.  To generate Receiver Operater Characteristic (ROC) curves we used the pROC (v1.8) R package.  A bonferroni adujusted p-value was used to assess differences between each study.

@@

In the final component of the analysis we performed power and sample number simulations for different effect sizes for each study.  This was completed for all the different components measured as part of first strategy.  We also calculated the actual sample size needed based on the effect size of each indivdiual study.  To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.

## Results:

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.

@@

**Power and Sample Size Estimates and Simulations**

@@

*******
##### Discussion:

@@

Third, although we may not have detected a signal between non-obese and obese indivdiuals our results indicate that we may not have been sufficiently powered to do so. However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.  Additionally, due to the very small signal another important question would be whether or not this difference is truly biologically meaningful.

@@

*******
##### Acknowledgements:

The authors would like to thank Nielson Baxter for his help and suggestions during the sequence analysis portion. We would also like to thank Shawn Whitefield for her help with making sure the manuscript adhered to the guidelines set out for a meta-analysis. We would also like to thank Rick Bushman, Hongzhe Li, and Pixu Shi in agreeing to release and helping to get a hold of the subject specific metadata from the Wu et al study.  Finally, we would like to thank all the authors who have made there data publicly available.  Without this initial forethought studies such as these would not be possible.

>>>>

## Introduction

Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014]. This number increases to approximately 35% in adults (aged 20 or older) and these statistics have seen little change since 2003 [@ogden_prevalence_2014]. Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013]. Recently, there has been increased interest in the role of the microbiome in modulating obesity [@brahe_can_2016; @dror_microbiota_2016]. If the microbiome does affect obesity status, then manipulating the microbiome could have a significant role in the future treatment of obesity and in helping to stem the current epidemic.

@@

One critique of these re-analysis studies is that they did not aggregate the results across studies to increase the effective sample size. It is possible that there are small associations within each study that would not be significant because the studies lacked sufficient statistical power. The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously. Furthermore, we sought to provide additional clarity on the possible relationship between community diversity, B/F, and community composition and obesity status and the ability to detect significant differences in these variables if they were actually present.

## Methods

***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1]. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified nine additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 9 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and one dataset was used (Baxter) because at the time of our analysis it was unpublished and enabled us to mitigate against publication biases associated with the bacterial microbiome and obesity. The nine studies are summarized in Tables 1 and 2.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, and number of OTUs. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. We also measured the Spearman correlation coefficient between each metric and the actual BMI value. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguish non-obese and obese individuals.  To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.  A Bonferroni-adjusted p-value was used to assess differences between each study.

@@

In the final component of the analysis we performed power and sample number simulations for different effect sizes for each study.  This was completed for all the different components measured as part of first strategy.  We also calculated the actual sample size needed based on the effect size of each individual study.  To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.

## Results

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a only 1 significant result for OTU richness. There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.

@@

***Power and Sample Size Estimates and Simulations***

@@

## Discussion:

@@

Third, although we may not have detected a signal between non-obese and obese individuals our results indicate that we may not have been sufficiently powered to do so. However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.  Additionally, due to the very small signal another important question would be whether or not this difference is truly biologically meaningful.

@@

## Acknowledgements

The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the manuscript. Finally, we would like to thank all the authors who have made there data publicly available.  Without this initial forethought studies such as these would not be possible.

### 2016-04-04 submission/Sze_Obesity_PeerJ_2016_Supp.Rmd
Light editing of text

The inital search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria.  The terms included in this criteria were that the manuscript had to have "Bacterial Microbiome" and "Obesity, BMI, bmi, obesity" in their manuscript criteria, it was not published more than 10 years ago, they were not review articles, and it contained research on humans only.  The below formula when put into PubMed should recapitulate our initial search on the website.

@@

From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequecning that used only clone libraries.  This ultimately excluded all but a total of 11 studies.

From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.  One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.  They also did not have obesity as part of their results in the actual published manuscript.  The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondance [@zeevi_personalized_2015].  A thrid study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]

@@

* PCR, qPCR, metagenomic sequencing, or RT-PCR used as main analysis Tool
* TRFLP or clone sequencing used to asses the bacterial community
* Utilization of 100bp or less single end reads for sequencing
* Sequencing Data not publicly available for download
* BMI not available and authors do not return correspondance
* Samples were not stool or feces
* Study contained children
* Study was a review

@@

```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

# Generate Funnel Plots for supplemental
par(mfrow=c(1,2))
funnel(res, main="A")
funnel(res2, main="B")

@@

```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

@@

```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/", dev = c("pdf", "png")}

>>>>

The initial search strategy included looking for all papers that initially fit under the below NCBI PubMed advanced search criteria.  The terms included in this criteria were that the manuscript had to have "Bacterial Microbiome" and "Obesity, BMI, bmi, obesity" in their manuscript criteria, it was not published more than 10 years ago, they were not review articles, and it contained research on humans only.  The below formula when put into PubMed should recapitulate our initial search on the website.

@@

From this total we browsed abstracts for mention of stool or feces examination, that did not involve children, was not a clinical trial for probiotics or other diet related treatments, did not only have participants with inflammatory bowel disease, the articles were in English, did not only use PCR, qPCR, or RT-PCR only for their analysis, and sequencing that used only clone libraries.  This ultimately excluded all but a total of 11 studies.

From this total of 12 studies the full text was reviewed for whether or not sequencing data was publicly available, BMI information (either categorical or continuous) was available in a supplement or, if it was not available, whether authors upon contact were willing to share this information or direct us to repositories that stored this specific information.  One study was excluded [@yatsunenko_human_2012] because it contained children and their sequencing of the 16S rRNA gene involved amplicons of only 100bp in length.  They also did not have obesity as part of their results in the actual published manuscript.  The second study was excluded because they did not have BMI information available and when contacted the authors never returned any correspondence [@zeevi_personalized_2015].  A third study was excluded since it did not use 16S rRNA gene sequencing for their bacterial microbiome analysis [@arumugam_enterotypes_2011]

@@

* PCR, qPCR, metagenomic sequencing, or RT-PCR used as main analysis Tool
* TRFLP or clone sequencing used to asses the bacterial community
* Utilization of 100bp or less single end reads for sequencing
* Sequencing Data not publicly available for download
* BMI not available and authors do not return correspondence
* Samples were not stool or feces
* Study contained children
* Study was a review

@@

```{r Figure_S4, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

# Generate Funnel Plots for supplemental
par(mfrow=c(1,2))
funnel(res, main="A")
funnel(res2, main="B")
```

@@

```{r Figure_S5, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

```{r Figure_S6, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

```{r Figure_S7, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

```{r Figure_S8, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

```{r Figure_S9, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}

@@

```{r Figure_S10, echo=FALSE, message=FALSE, warning=FALSE, fig.path="results/figures/"}
